---
title: "ANOVA for Journalists: Analyzing College Admissions Test Scores"
date: 11-3-2025
---
## Name: Diya Shah


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this exercise, we'll explore a common journalism question: *Are there significant differences in standardized test scores across different types of colleges?*

We'll use ANOVA (Analysis of Variance) to determine if the observed differences in SAT scores across three college types (Public, Private, and Community College) represent real patterns or just random variation.

## Task 1: Load Libraries and Create Sample Data

```{r}
library(tidyverse)
```

```{r create-data}
# Set seed for reproducibility
set.seed(123)

# Create simulated SAT scores for admitted students across three college types
# Scale is 400-1600 for total SAT
# rnorm generates random numbers, in this case, 100 numbers
public_college <- rnorm(100, mean = 1350, sd = 120)
private_college <- rnorm(100, mean = 1450, sd = 100)
community_college <- rnorm(100, mean = 1050, sd = 150)

# Combine into a single dataframe
admissions_data <- data.frame(
  college_type = factor(rep(c("Public", "Private", "Community"), each = 100)),
  sat_score = c(public_college, private_college, community_college)
)

# Ensure SAT scores are within realistic range (400-1600)
admissions_data$sat_score <- pmin(pmax(admissions_data$sat_score, 400), 1600)

# Quick look at the data
head(admissions_data)
```

## Task 2: Calculate Basic Statistics

```{r summary-stats}
# Summary statistics by college type
admission_summary <- admissions_data |>
  group_by(college_type) |>
  summarize(
    mean_score = mean(sat_score),
    median_score = median(sat_score),
    sd_score = sd(sat_score),
    min_score = min(sat_score),
    max_score = max(sat_score)
  )

print(admission_summary)
```

Context: Fall 2024 midpoint SAT for University of Maryland was 1,465.


**Task 3** Describe the data

Write a couple of sentences describing the differences in SAT scores between types of colleges, using the descriptive statistics we just generated.


**Task 3: Private universities have the highest average SAT scores with the lowest variation, which means that a lot of students have SAT scores that are around the mean (1435). Community colleges had the largest range of SAT scores, with the lowest being 786 and the highest 1393, giving it a range of 607, compared to the range of public school scores which is 528 and private schools which is 356. Community college scores also have the most amount of variation. **


## Task 4: Visualize the Data

Let's create visualizations to see the patterns:

```{r visualize-distributions}
# Create density plot showing overlapping distributions
ggplot(admissions_data, aes(x = sat_score, fill = college_type)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "SAT Score Distributions by College Type",
    subtitle = "Are these differences statistically significant?",
    x = "SAT Score (400-1600 scale)",
    y = "Density",
    fill = "College Type"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

```{r boxplot}
# Create boxplot for another view
ggplot(admissions_data, aes(x = college_type, y = sat_score, fill = college_type)) +
  geom_boxplot() +
  labs(
    title = "SAT Scores by College Type",
    x = "College Type",
    y = "SAT Score"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

## Task 5: Perform ANOVA Test

```{r anova-test}
# Run one-way ANOVA
anova_result <- aov(sat_score ~ college_type, data = admissions_data)
summary(anova_result)
```

The F-value is huge (284.3). The F-value measures the ratio of between-group variance to within-group variance, and this high value suggests the differences between college types far outweigh the variation within each type. The p-value is very low, indicating the differences between college types are statistically significant at any reasonable significance level. The three asterisks (***) highlight that this is significant at the 0.001 level. So we can say that with confidence that there are real differences in SAT scores between the three types of colleges.

How do we know the F-value is huge?
We can calculate the critical value for an F statistic by using the qf() function in R

```{r}
# Your ANOVA has:
# df1 (between groups) = 2
# df2 (within groups/residuals) = 297
# Common alpha levels: 0.05, 0.01, 0.001

# Critical values at different significance levels:
critical_f_05 <- qf(0.95, df1 = 2, df2 = 297)
critical_f_01 <- qf(0.99, df1 = 2, df2 = 297)
critical_f_001 <- qf(0.999, df1 = 2, df2 = 297)

cat("Here are the Critical F-values:\n")
cat("alpha of α = 0.05: ", round(critical_f_05, 3), "\n")
cat("alpha of α = 0.01: ", round(critical_f_01, 3), "\n")
cat("alpha ofα = 0.001:", round(critical_f_001, 3), "\n")
cat("The F-statistic is: 284.3\n")
cat("Ratio to most extreme critical value (α=0.001) to the F statistic:", round(284.3/critical_f_001, 1), "x\n")
```


## Task 5: Compare the differences between specific pairs

```{r post-hoc}
# Tukey's HSD test to see which specific pairs of college types differ
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

The Tukey multiple comparisons test shows highly significant differences in SAT scores between all college types:

* Private vs. Community: Private colleges have SAT scores that are 366.96 points higher than community colleges on average (p < 0.0001)
* Public vs. Community: Public colleges have SAT scores that are 292.55 points higher than community colleges on average (p < 0.0001)
* Public vs. Private: Public colleges have SAT scores that are 74.40 points lower than private colleges on average (p = 0.000021)

All comparisons show extremely small p-values, indicating these differences are statistically significant at any conventional significance level. The 95% confidence intervals (shown in the "lwr" and "upr" columns) do not include zero for any comparison, further confirming these differences are statistically significant.

## Task 6: Report the Findings

Based on our analysis, we can now answer:

1. Is there a significant difference in SAT scores across college types?
2. Which specific college types differ from each other?
3. How large are these differences?

Answer those questions, providing evidence for your answers, and then write a headline and a lead paragraph reporting these findings using active voice and focusing on the differences across types.

1. There is a significant difference in SAT scores across college types because all of the p-values are all extremely low (less than 0.001 which demonstrates that it is statistically significant). 

2. The Public and Private colleges specifically statistically differ from each other based on the tests that we conducted. Based on the Tukey test that we did, the difference between SAT scores from public and private colleges is -74.4, which shows that there is a large statistical difference. While the differences between private and community and public and community are still statisically significance, the difference isn't as big as it between public and private.

3. The difference between public and private is the largest difference (-74.4). The difference between public and community is 292.55 and the difference between private and community is 366.96. 

Headline: Private schools maintain the highest SAT scores, in comparison to public and community colleges, data shows 

SAT scores vary significantly between different college types, with private colleges leading with the highest SAT scores by wide margins. According to statistical testing, private colleges have SAT scores that average nearly 367 points higher than community colleges and about 74 points higher than public colleges. The differences in SAT scores between the types of colleges are statistically significant All differences are statistically significant, as based on testing, they all have extremely small p-values, indicating that the gaps in average SAT scores between college types are not due to chance. 
