---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2020/pop2020 *1000, property_rate_2020 = property2020/pop2020 * 1000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The two cities with the highest violent crime rate in 2019 were Baltimore and then Ocean City, those two were had the highest violent crime rate in 2020 as well, except Ocean City's was higher than Baltimore in 2020. The violent crime rate in Ocean City increased by over 10 points between 2019 and 2020 (13.1% to 24%), while Baltimore's violent crime rate decreased by around 2 points (18.3% to 16%). Ocean City had the highest property crime rates in both 2019 (103.7%) and 2020 (121.3%). Based on these results, I feel like the story could be about questioning why Ocean City has such high crime rates -- does it have to do with the fact that its a beach town? lots of tourism? Does crime spike during the summer and then decrease during the off season? 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate (violent_pct_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100, property_pct_change = (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100) |> select(city, pop2020, pop2019, violent2019, violent2020) # <-- this is how you can organize columns (use select and then list the column names in the order you want to see it)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Based on result of the percent change calculations, the percent changes seemed to be pretty drastic. It was tough to distinguish or figure out a pattern, so I think that the outliers are more interesting. Based on the data I think the story would be questioning why there are such huge changes in just one year. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Violent crime in Ocean City jumped from 2019 to 2020, giving the beach town the highest violent crime rate in the state of Maryland. 

Crime data shows that Ocean City's violent crime rate rose from 13.1% to 24% in just one year, even as Baltimore, a city that has long been considered Maryland's most dangerous, saw a slight decline. Ocean City also recorded the highest property crime rates in both years, calling into question whether Ocean City's tourism industry has played a significant role in the crime rates going up.  

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

## task 10: hw 5 percentiles

```{r}
md_cities_percentiles <- md_cities_rates |>
mutate(percentile = ntile(violent_rate_2020, 100))
```

1. 10 cities are above the 90th percentile. 
2. I think that the results from calculating the IQR for violent crime rates in 2020 wouldn't change my story idea. The IQR shows the middle 50% of a dataset, and the IQR, being 4.2, shows that Ocean City's violent crime rate is an outlier compared to most of the other cities, supporting my original story idea about how Ocean City's violent crime increase is newsworthy and higher than what you would expect from the typical range of maryland cities. 

```{r}
md_cities_percentiles |>
summarize(IQR_violent_rate_2020 = IQR(violent_rate_2020))
```